---
title: "Line Chart Refresher: WMATA Ridership"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Quick Review: Making a Line Chart

Let's do a quick refresher on line charts using WMATA ridership data, since a line chart will be part of your homework. This is daily bus and rail data from January 2024 through February 2025 (you may recall it from the midterm). We'll aggregate it to weekly totals so the chart is easier to read, then make a line chart.

### Task 1: Load the tidyverse and the data

```{r}
library(tidyverse)
library(lubridate)

wmata <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")

head(wmata)
```

### Task 2: Make weekly totals

Daily data is noisy — weekends and holidays create big dips that make trends hard to see. Let's aggregate to weekly totals using `floor_date()` to group days into weeks, then `summarize()` to add them up.

```{r}
weekly_wmata <- wmata |>
  mutate(week = floor_date(date, "week")) |>
  group_by(week) |>
  summarize(
    weekly_total = sum(total),
    days = n()
  ) |>
  filter(days == 7)  # Keep only complete weeks

head(weekly_wmata)
```

### Task 3: Make a line chart

Create a line chart of weekly ridership over time.

```{r}
weekly_wmata |>
  ggplot(aes(x = week, y = weekly_total)) +
  geom_line(color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Weekly WMATA Ridership (2024-2025)",
    x = "Week",
    y = "Total Riders"
  ) +
  theme_minimal()
```

**Answer this question:** Describe the trend you see. Is ridership going up, down, or staying flat? What are the big dips, and what might explain them? What would you want to know before writing a story about Metro weekly ridership trends?

The trend line shows a very slight downward trend in WMATA ridership from January 2024 to January 2025. The really big dips occur around major holidays where people are staying home or traveling to family members without using public transit, (in this case, the big dips seem to occur at Thanksgiving, Christmas, New Year's and MLK Day). Before writing a story about WMATA's ridership trends, I would like to see the data from January 2023 to January 2024, does this slight downward dip always occur because of the major holidays late in the year, allowing the numbers to bounce back early in the start of the next?
