---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Andy Parr"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

I'm much more confident in this dataset than the previous one because there isn't the giants red flag of having eight different schools with 100% as their FARMS percent. That said, there are nine schools that have a percentage of exactly 4%, which seems pretty strange. Also the overall range is huge, going from 4% to 96%, which feels like a giant range for something that feels like it should be grouped a little bit more.


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

The first thing that stands out is the difference in the strengths of the charts. This one, (Maryland), has its line of best fit at a much sharper angle than the other scatterplot (Pennsylvania). The next thing is that this one actually has a negative relationship, unlike the Pennsylvania plot that had a slightly positive slope to its line. This data has a much cleaner spread too, hitting perentages all over the scatterplot unlike the Pennsylvania chart, which shoots from a little over 50% and straight to 100% values.

Maybe they're different because of factors like income, amount of schools, faulty data, etc. Very few things in life are truly linear, so even though they have the same name and general geography, there's still a lot of other factors in play.


### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

The p-value is very near zero (2.2e16), the adjusted r-squared is about 0.67 and the coefficient is about -0.60

These three pieces of information say that there is a statistically significant change when analyzing FARMS percent and math scores AND the relationship is a real thing, the line of best fit isn't too much of a stretch. It has a defenseable claim that schools with a higher percentage of students on FARMS tend to perform below expected levels for math testing. This is a model worth using because it's statistically significant and has a strong line of best fit. Unlike the pretty much unusable Pennsylvania model, which was just barely within the p-value threshold but struggled in finding a true relationship between variables.


### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

The largest positive residual belongs to Wayside Elementary (about 56.46) and the largest negative residual is Stedwick Elementary (about -55.05).

Wayside Elementary is zip code 20854. The area is mostly white with 17% percent of people aged 10-19 and 16% aged 50-59. The median household income is quite literally JUST over 250K and only 2.6% of residents live below the poverty line. I would say that can help explain why the school is overperforming, it's in a really nice area and probably has access to better resources for its students. My first question to the principal would be "What strategies do you use with your teachers to make sure your students understand key concepts?" Based on that answer I can see whether or not they use a bunch of extra resources that might not be possible for other schools in the county.

Stedwick Elementary is zip code 20886. It is majority hispanic with high black and white populations as well. It's on the younger side, 15% of people are 0-9 and 16% are 30-39. The median household income is on the lower side, about 98.5K with 12.8% of residents living below the poverty line. My question for the principal is "If you could add one new thing to help your students, what's at the top of the list?" Similar to the last one, see if resources are the main thing in play here.

### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

The Maryland data supports the school board member's claim with a statistically significant change and a defenseable model in favor of the statement. This is a much better model than the one for Montgomery County Pennsylvania, which while statistically significant, lacked a defendable relationship between the variables. If I asked the school board member for comment, I would ask what their analysis looked like, (see if they ran tests similar to this or used a different process) and their plans for how to address the issue of lower standardized math scores in schools with higher rates of FARMS. Lastly, the school I'd want to see first is Wayside Elementary. I want to get a feel for what that place is like and if it seems noticably nicer than the average school, if it's obvious they have a resource advantage, etc.
