---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
mandms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset

```{r}
sample_5 <- mandms |> sample_n(5)
```


### Task 4: Calculate the mean age and standard deviation of the sample, using either the blue or white number of M&Ms

#Sample has red and black

```{r}
sample_5 |> summarise(mean_black = mean(black), sd_black = sd(black))
```


### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4

```{r}
mandms |> summarise(mean_black_pop = mean(black), sd_black_pop = sd(black))
```


### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

The means are pretty close to each other, (73.28 population and 69 sample), but the standard deviations are VERY close to each other (15.75 population and 15.43 sample). Given that there's only 100 samples in the total population, it's not that surprising that the results are so close.

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

For random sampling to be accurate, the larger the sample size, the better. I will admit I got lucky with my random sample being so close to the mean and standard deviation of the population, but as mentioned before, this is a smaller population than we've been dealing with so far this semester. There were only 100 possible samples, so picking out five random ones means it'll likely be closer to the mean and standard deviation than if there were 1,000 or 10,000 samples to pick from

While my data ended up being close to the true population values, if I had gotten 20 samples instead of five, it likely would've been even closer to the population values.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
