---
title: "final_project_clean"
output: html_document
author: Andy Parr
---
Step 1: Load up the libraries/data. Also do this filter to make sure you're using the correct rows for the goals data. If other rows are included then it will overestimate each year's goal scoring.
```{r}
#loading libraries

library(tidyverse)
library(janitor)
library(corrplot)

#loading data

better_goals_data <- read_csv("teams_2008_to_2024.csv") |> clean_names()

better_goals_data <- better_goals_data |> filter(str_detect(situation, "all")) # <-- This step is CRUCIAL so that all the stats are in one place instead of separating each team into four different rows for each season (data source: https://moneypuck.com/data.htm?utm_)
```
Step 2: Select the right columns from the data set to create a new dataframe that includes average goals, shots and saves, plus qualifiers for the major rule changes. Additionally, this step finds the lm models for each variable and the rule changes.

Based on the p-values that this code calculated, it looks like the 2018 rule change may have had some actual effects on goal scoring. The next step will visualize what the data found.
```{r}
# Now using the code you helped me fix, thanks!

league_goals <- better_goals_data |>
  group_by(season) |>
  summarize(goals_per_game = sum(goals_for) / sum(games_played)) |>
  mutate(
    post_2006 = season >= 2006,   # post-lockout equipment cut
    post_2013 = season >= 2013,   # trapezoid expansion
    post_2018 = season >= 2018    # chest/arm + pants reductions
  )

lm(goals_per_game ~ season + post_2006 + post_2013 + post_2018, data = league_goals) |>
  summary()

league_shots <- better_goals_data |>
  group_by(season) |>
  summarize(shots_per_game = sum(shots_on_goal_for) / sum(games_played)) |>
  mutate(
    post_2006_shots = season >= 2006,   # post-lockout equipment cut
    post_2013_shots = season >= 2013,   # trapezoid expansion
    post_2018_shots = season >= 2018    # chest/arm + pants reductions
  )

lm(shots_per_game ~ season + post_2006_shots + post_2013_shots + post_2018_shots, data = league_shots) |>
  summary()

league_save <- better_goals_data |>
  group_by(season) |>
  summarize(save_per_game = sum(shots_on_goal_against - goals_against) / sum(games_played)) |>
  mutate(
    post_2006_sa = season >= 2006,   # post-lockout equipment cut
    post_2013_sa = season >= 2013,   # trapezoid expansion
    post_2018_sa = season >= 2018    # chest/arm + pants reductions
  )

lm(save_per_game ~ season + post_2006_sa + post_2013_sa + post_2018_sa, data = league_save) |>
  summary()
```
Step 3: Here's that visualization! I created a scatterplot with a a line of best fit that highlighted the 2018 season. I also compared it to league-wide shots per season to show that shot volume stayed the same during the time frame.
```{r}
scoring_bump_2018 <- league_goals |> 
  ggplot() +
  geom_point(aes(x=season, y=goals_per_game)) +
  geom_smooth(aes(x=season, y=goals_per_game), method="lm") +
  geom_vline(xintercept = 2018, color = "red", size = 1, linetype = "solid") +
  labs(
    x = "Year",
    y = "Goals per Game",
    title = "NHL League-Wide Goal Scoring per Year"
  )
scoring_bump_2018

year_by_year_shots <- league_shots |> 
  ggplot() +
  geom_point(aes(x=season, y=shots_per_game)) +
  geom_smooth(aes(x=season, y=shots_per_game), method="lm") +
  geom_vline(xintercept = 2018, color = "red", size = 1, linetype = "solid") +
  labs(
    x = "Year",
    y = "Shots per Game",
    title = "NHL League-Wide Shots on Goal per Year"
  )
year_by_year_shots
```

Step 4: Join all those new dataframes to do some math, check if there's something bigger going on here beyond the initial lm models.
```{r}
league_combined <- league_goals |>
  left_join(league_shots |> select(season, shots_per_game), by = "season") |>
  left_join(league_save |> select(season, save_per_game), by = "season")
```
Step 5: Perform a few different multiple regression tests to see which has the strongest relationship and visualize.

What's weird here is that Model 1 and Model 2 both came out with a perfect adjusted r-squared, which likely means that something is off with this particular analysis. For what it's worth, models 3 and 4 (designed to test whether or not the rule changes were the major factor), also had a high adjusted r-squared, just not a perfect one.

Due to these weird results, this test feels like it has some holes and will not factor into my final story pitch, but it was worth a try.
```{r}
# Model 1 (Full model)

model1 <- lm(goals_per_game ~ post_2013 + post_2018 + shots_per_game + save_per_game, data = league_combined)
summary(model1)

# Model 2
model2 <- lm(goals_per_game ~ shots_per_game + save_per_game, data = league_combined)
summary(model2)

# Model 3
model3 <- lm(goals_per_game ~ post_2013 + post_2018, data = league_combined)
summary(model3)

# Model 4
model4 <- lm(goals_per_game ~ post_2013 + post_2018 + save_per_game, data = league_combined)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(model1)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(model1)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```
Step 6: Analysis

Following all these code adjustments, regression analysis and significance tests, the NHL's effort to increase goal scoring through rule changes worked in 2018 when the league forced goalies to use smaller, less bulky equipment. Goals went up by about .217 per game across the league starting in that post-rule change era and the p-value is below 0.05 (statistically significant). Additionally, the shots and saves per game did NOT see a significant change during the same time frame, indicating that there were not significant changes in the way teams were playing on the ice, but rather the league made it easier to score and a greater percentage of shots started beating goalies after the rule changes.

In other words, (without math), this analysis indicates that scoring increased following the 2018 goalie equipment regulation changes because it made team's shots more efficient. There was no real change in the shot volume in the same time.

STORY PITCH: The NHL's goalie pad adjustment in 2018 increased league-wide scoring by a little more than 0.21 goals per game. In the same time, there were no significant changes in shots taken, meaning teams became more efficient with scoring, this was not a result of higher shot volume.

ISSUES: This was addressed earlier, but I had problems with the multiple regression in steps 4 and 5. I simply couldn't trust those test results when two of the models came back as having a perfect correlation.

REPORTING PLAN: Check with NHL sources to see if they intentionally tried to increase goal scoring with any of these rule changes. Ask whether or not they stopped following the 2018 season because they got such a big spike. Also check in with players who were in the league before and after these adjustments and see if they felt any big differences.
